\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-06-21 main.tex
merge all the Rmds into a single oneg

\author{
H. Sherry Zhang\\Monash University \And Dianne Cook\\Monash University
\AND Ursula Laa\\University of Natural Resources and Life Sciences
\AND Nicolas Langrené\\BNU-HKBU United International College
\AND Patricia Menéndez\\Monash University \AND
}
\title{\pkg{cubble}: An R Package for Structuring Spatio-temporal Data}

\Plainauthor{H. Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas
Langrené, Patricia Menéndez}
\Plaintitle{cubble: An R Package for Structuring Spatio-temporal Data}

\Abstract{
Spatio-temporal data refer to measurements taken across space and time.
In practice, spatio-temporal data can be decomposed into a spatial and
temporal component: at one time, we would select a spatial location and
inspect the temporal trend; at other time, we might select one or
multiple time value(s) and explore the spatial distribution. Ideally, we
could make multiple maps and multiple time series to explore these
together, however, doing all of these actions is complicated when data
arrive fragmented in multiple objects. To make it easy to do all these
tasks, ideally spatial and temporal variables are in a single data
object that we can slice and dice in different ways to conduct different
visualisations. This work suggests a new data structure, \code{cubble},
that uses a nested form and a long form to organise spatial and temporal
variables in a single data object. The new data structure ensures that
data in the two forms are synchronised while been flexible to explore
the two aspect of spatio-temporal data. It also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics, and performing
spatio-temporal transformation. The paper will demonstrate \pkg{cubble}
with examples using Australian climate weather stations, river level
data, and climate reanalysis (ERA5) data.
}

@@

      Ursula Laa\\
    University of Natural Resources and Life Sciences\\
    Gregor-Mendel-Straße 33, 1180 Wien, Austria\\
  E-mail: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}\\

@@

\usepackage{amsmath} \usepackage{array} \usepackage{xcolor}

@@

\newpage

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variables. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

@@

\caption[An illustration of different spatio-temporal data layouts]{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

While incorporating both spatial and temporal information enriches the
data in the analysis, spatio-temporal data can sometimes be hard to work
with given its two observational units in the variables. The single
table format in Figure \ref{fig:illu-input} is convenient for
spatio-temporal operations that involve both types of variables,
however, it makes duplicates of spatial variables and is not ideal when
only looking at the spatial aspect of the data. The two tables format
allows separate operations on spatial and temporal variables while
analysts need to work with two data objects simultaneously and join them
together into the single table format for spatio-temporal operations.
Neither of the two formats is desirable at the same time for spatial,
temporal, and spatio-temporal operations. This issue implies analysts
need to constantly think about the data format and arrange variables
into the proper shape before the actual operations, which can be a
laborious step when switching between the two formats back and forth
several times during the analysis.

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal classes it is built on have
now been gradually replaced by more recent implementations. The
\pkg{stars} \citep{stars} package uses an array to wrangle
spatio-temporal data, but the dense array it uses can quickly exhaust
the memory for data in the case of irregular spatial grid. While there
has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With this new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use of \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.

\hypertarget{conceptual-framework-spatio-temporal-cube}{%
\section{Conceptual framework: spatio-temporal
cube}\label{conceptual-framework-spatio-temporal-cube}}

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by \citet{lu_multidimensional_2018} has discussed the
suitability of this representation through array algebra and practical
spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the
operation can be done on the data for visualisation. A paper by
\citet{bach_review_2014} has reviewed the temporal data visualisation
based on space-time cube operations. Notice that the term space-time
cube in their article ``does not need to involve spatial data'', but
referring to ``an abstract 2D substrate that is used to visualize data
at a specific time''. Despite its main focus is on temporal data, the
mindset of abstracting out data representation to construct
visualisation is applicable to our spatio-temporal data manipulation and
visualisation.

\hypertarget{existing-work-and-new-challenges}{%
\section{Existing work and new
challenges}\label{existing-work-and-new-challenges}}

To represent spatio-temporal data, \citet{spacetime} proposes four
layouts: full grid layout (STF), sparse grid layout (STS), irregular
layout (STI), and trajectory (STT). These layouts are implemented in the
\pkg{spacetime} package with underlying class \pkg{sp} \citep{sp} and
\pkg{xts} \citep{xts} for handling spatial and temporal variables. While
these layouts and classes are still being used and well-maintained
today, they are not compatible with the \pkg{tidyverse} ecosystem
\citep{tidyverse}, which has a huge influence on data analysis in
\proglang{R}, including in the spatial and temporal communities. The
package \pkg{sf} \citep{sf} uses data frame objects to represent spatial
data with an additional \code{geometry} column dedicating to feature
geometries \citep{jr_herring_opengis_2011} (e.g.~POINT, MULTIPOINT,
POLYGON, MULTIPOLYGON etc). For temporal data, \pkg{tsibble}
\citep{tsibble} builds on top of the \pkg{tibble} \citep{tibble} object
and explicitly includes date and time in columns as variables, rather
than data attributes.

Underlying tidyverse is the concept of tidy data \citep{tidydata}, which
considers how variables should be structured for analysis. In the three
formats (time-wide, space-wide, and long) summarised in the
\citet{spacetime} paper, only the long form is considered as tidy. The
main advantages of it over the time-wide and space-wide format is that
the latter treat either time or location as variable names (Section 3.1
in \citet{tidydata}) rather than values of variables, making them
difficult to be wrangled. One problem with the long format is that it
can be inefficient to store feature geometries, especially for large
MULTIPOLYGON over long time period. This motivates a new tidyverse
compatible data object to work with spatio-temporal data.

\textbf{two challenages: multiple data sources + large spatio-temporal
object} In additional to the shift into more tidyverse workflow for
working with data, spatio-temporal data have some unique challenges. The
first one being the split of variables in multiple tables. Often the
time, spatio-temporal data is not presented in a single data object, but
split into fragmented pieces from different sources. For example, map
data recorded the shape of a collection of area of interest is provided
by the government website; the geo-scientific data is recorded with
longitude and latitude coordinate of locations in the area; and temporal
variables indexed by both location and time need to be further queried
using location data. For spatio-temporal data analysis, merging all
these sources can be challenging and joining these tables can go wrong
because of the slightest unmatched of the linking variable from
different sources.

Another challenge when analysing spatio-temporal data is on the simple
feature representation of spatial data. While the simple feature is able
to handle large and complex spatial geometry and perform precise spatial
operations, data analysts usually do not have a large expectation on the
map accuracy and would prefer a simplified map with quick rendering for
visualisation. Also, the feature geometry column in \pkg{sf} wraps the
spatial coordinates in a list column, sometimes, analysts may like to
unpack them into longitude and latitude to work with. A smooth
transition with the coordinate columns and and simple feature geometry
column would satisfy analysts from both ends.

With some of the aforementioned issues recognised, the \pkg{stars}
package has been a new implementation to handle spatio-temporal data
conceptually as data cubes and implemented as arrays. The package
depends on the \pkg{sf} package and has allowed more access to tidyverse
verbs in the workflow. However, for vector spatio-temporal data, major
difficulties to work with \pkg{stars} is first to cast the variables
from multiple tables into a star object and then keep the operations in
the high-dimensional cube. Given that most table mentioned before
collect variables in a 2D table, analysts could benefit from a structure
to allow them work in 2D tables.

\hypertarget{the-cubble-package}{%
\section{The cubble package}\label{the-cubble-package}}

This section will present two formats that cubble uses to arrange
spatio-temporal data. The main cubble functions will then be introduced
to illustrate how to work with these two formats in short examples.
Lastly, a subsection will be dedicated to how existing packages, in
spatial and temporal analysis, fit in with cubble.

In cubble, a single data object can be arranged into two forms: nested
form and long form. Figure \ref{fig:illu-cubble} sketches the two forms
with the associated attributes. The decision on which form to use in
data wrangling is output-oriented, meaning analysts need to first think
about whether the output of an operation is identified only by the
spatial identifier, or a combination of spatial and temporal identifier.
The nested cubble is suitable for working with operations that are only
identified by the site and this type of operation can be a pure
manipulation of time invariant variables, or an operation that
summarises time varying variables into site. Underneath the nested form,
a cubble is built from a \code{rowwise_df} class where each site forms a
separate group. This structure simplifies the calculation that involves
temporal variables by avoiding the use of \code{purrr::map()} syntax
when working with list-column.

For those operations whose output involves both a spatial and temporal
identifier, long form should be used. The long form uses a
\code{grouped_df} class to forms all the time of a site as a group. Time
invariant variables are stored separately as an special attribute of the
long cubble. This design avoids repeating the spatial variables at each
time stamp while not dropping information from spatial variables.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.002}

@@

\caption{An illustration of the nested and long form in cubble. The nested form defines each site in a row and nests the time varying variables into a single column \code{ts}. The long form cubble uses \code{id} and \code{t} to identify each row and stores the time invariant variables as an attribute, \code{spatial}.}\label{fig:illu-cubble}
\end{figure}
\end{CodeChunk}

\hypertarget{create-a-cubble-in-the-nested-form}{%
\subsection{Create a cubble in the nested
form}\label{create-a-cubble-in-the-nested-form}}

To use functions in the \texttt{cubble} package, an analyst will first
need to turn the data object into a \code{cubble} class.
\code{as_cubble()} does this through supplying the three key components:
\code{key} as the spatial identifier; \code{index} as the temporal
identifier; and a vector of \code{coords} in the order of longitude and
latitude. The use of \code{key} and \code{index} follows the design in
the \pkg{tsibble} package and the cubble created by default is in the
nested form.

Before the formal examples in Section \ref{examples}, each function
introduced in this section will be accompanied by a small example. The
data used for the small examples is a subset of a larger climate weather
station data used in the formal examples and has been simplified to
include only five weather stations. It contains spatial information of
each station: station id, latitude, longitude, elevation, station name
and World Meteorology Organisation ID and also daily temporal
information of date, maximum and minimum temperature and precipitation
for 2020. \code{climate_flat} stores this data in format 1) in Figure
\ref{fig:illu-input} and the code below creates a cubble out of
\code{climate_flat} with \code{id} as the key, \code{date} as the index,
and \code{c(long, lat)} as the coordinates:

>>>>

\author{
H. Sherry Zhang\\Monash University \And Dianne Cook\\Monash
University \And Ursula Laa\\Universität für Bodenkultur Wiens
\AND Nicolas Langrené\\BNU-HKBU United International
College \And Patricia Menéndez\\Monash University
}
\title{\pkg{cubble}: An \proglang{R} Package for Organizing and
Wrangling Multivariate Spatio-temporal Data}

\Plainauthor{H. Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas
Langrené, Patricia Menéndez}
\Plaintitle{cubble: An R Package for Organizing and Wrangling
Multivariate Spatio-temporal Data}

\Abstract{
Multivariate spatio-temporal data refers to multiple measurements taken
across space and time. For many analyses, spatial and time components
can be separately studied: for example, to explore the temporal trend of
one variable for a single spatial location, or to model the spatial
distribution of one variable at a given time. However for some studies,
it is important to analyze different aspects of the spatio-temporal data
simultaneously, like for instance, temporal trends of multiple variables
across locations. In order to facilitate the study of different portions
or combinations of spatio-temporal data, we introduce a new class,
\code{cubble}, with a suite of functions enabling easy slicing and
dicing on different spatio-temporal components. The proposed
\code{cubble} class ensures that all the components of the data are easy
to access and manipulate while providing flexibility for data analysis.
In addition, the \pkg{cubble} package facilitates visual and numerical
explorations of the data while easing data wrangling and modelling. The
\code{cubble} class and the functions provided in the \pkg{cubble}
\proglang{R} package equip users with the capability to handle
hierarchical spatial and temporal structures. The \code{cubble} class
and the tools implemented in the package are illustrated with different
examples of Australian climate data.
}

@@

      Ursula Laa\\
    Universität für Bodenkultur Wiens\\
    Gregor-Mendel-Straße 33, 1180 Wien, Austria\\
  E-mail: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}\\

@@

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

\usepackage{amsmath} \usepackage{array}

@@

@@

Spatio-temporal data has a spatial component referring to the location
of each observation and a temporal component that is recorded at regular
or irregular time intervals. It may also include multiple variables
measured at each spatial and temporal values. With spatio-temporal data,
one can fix the time to explore the spatial features of the data, fix
the spatial location/s to explore temporal aspects, or dynamically
explore the space and time simultaneously.

In order to computationally explore the spatial, temporal and
spatio-temporal faces of such data, the data needs to be stored and
represented under a specific data object that allows the user to query,
group and dissect all the data faces.

The Comprehensive \proglang{R} Archive Network (CRAN) task view
SpatioTemporal \citep{ctvspatiotemporal} gathers information about
\proglang{R} packages designed for spatio-temporal data and it has a
section on \emph{Representing data} that lists existing spatio-temporal
data representations used in \proglang{R}. Among them, \citet{spacetime}
summarises spatio-temporal data into three forms: time-wide, space-wide,
and long formats. The associated package \pkg{spacetime}
\citep{spacetime} implements four spatio-temporal layouts (full grid,
sparse grid, irregular, and trajectory) to handle different space and
time combinations. The package \pkg{stars} \citep{stars} has a new
implementation to use dense arrays to represent spatio-temporal cubes.
It also interfaces with the package \pkg{sf} \citep{sf}, commonly used
for wrangling spatial data, and the \pkg{tidyverse} \citep{tidyverse}
suite for general data wrangling and visualization in \proglang{R}.

Still, the data representation for spatio-temporal data can be further
extended and there are two reasons for this. Firstly, the raw data
sourced in the wild is less often presented in any one of the layouts
above, and fitting the raw data into a data object can sometimes be
difficult. More often, spatio-temporal data are collected in separate 2D
tables and analysts need to assemble them into a whole piece before
exploring the data. Examples of components of spatio-temporal data can
be 1) areal data recording the shape of a collection of areas of
interest; 2) geostatistical data storing the longitude and latitude
coordinates of locations, typically also with other metadata related to
the location, and; 3) temporal data of each location across time.

The other reason is about tidy data concepts \citep{tidydata} and how
they should be applied to spatio-temporal data. According to the tidy
data principles, data should be structured into 1) one row per
observation, 2) one column per variable, and 3) one type of data per
table. The long form data is preferred over wide data form given the
downstream packages such as \pkg{dplyr} \citep{dplyr} and \pkg{ggplot2}
\citep{ggplot2} for data wrangling and visualization. However, the long
form can be inefficient to store feature geometries, especially for
large multipolygons for hourly, daily or sub-daily periods over years,
which are extensively collected and handled, for example in time series
analysis. This poses the question of how to arrange spatial and temporal
variables in a way that would make data wrangling, visualizing and
analyzing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, which
addresses the two issues mentioned above. In the package, a new class,
also called cubble, is proposed to organize spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately or combined while being kept synchronized. Among the
four spacetime layouts in \citet{spacetime}, the \code{cubble} class can
be applied to full grid, sparse grid, or irregular, but not trajectory,
which is outside the scope of this work. The software is available from
the Comprehensive \proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=cubble}.

The rest of the paper is organized as follows: Section @ref(cubble)
presents the main design and functionality of the \pkg{cubble} package.
Section @ref(others) explains how the \pkg{cubble} package deals with
more advanced considerations, including data with hierarchical
structure, data matching and how the package fits with existing static
and interactive visualization tools. Moreover we also illustrate how the
\pkg{cubble} package deals with spatio-temporal data transformations.
Section @ref(examples) uses Australian weather station data and river
level data as examples to demonstrate the use of the package. An example
of how the \pkg{cubble} package handles Network Common Data Form
(NetCDF) data is also provided. Section @ref(conclude) discuss the paper
contributions and future directions.

\hypertarget{cubble}{%
\section{The cubble package}\label{cubble}}

\hypertarget{object}{%
\subsection{The cubble object}\label{object}}

Spatio-temporal data can encompass data with various spatial and
temporal characteristics and different data require different structures
for wrangling and analysis: climate weather stations record data at
fixed point location but may suffer from potential temporal data quality
issue such as missing data for certain days. GPS data tracks unique
point locations at different timestamps and is represented as
trajectories. Satellite imageries capture snapshots of landscapes at
selected time and is commonly structured as raster data. The
spatio-temporal data cubble address are those collected at unique fixed
locations while allowing for irregularity in the temporal dimension,
such as the weather station data. This corresponds to the full
space-time and sparse space-time layouts described in the spacetime
paper \citep{spacetime} and reproduced in Figure \ref{fig:layout}.

@@

{\centering \includegraphics{/Users/hzha400/Documents/research/paper-cubble/figures/layout-1}

@@

\caption[The full space-time and sparse space-time layouts handled by the cubble package]{The full space-time and sparse space-time layouts handled by the cubble package. Reproduced from Figure 1 in the spacetime paper \citep{spacetime}.}(\#fig:layout)
\end{figure}
\end{CodeChunk}

The cubble class is an S3 class built on tibble that allows the
spatio-temporal data to be wrangled in two forms: a nested form (spatial
cubble) and a long form (temporal cubble). It consists of two
subclasses:

\begin{itemize}
\tightlist
\item
  a spatial cubble with class
  \texttt{c("spatial\_cubble\_df",\ "cubble\_df")}
\item
  a emporal cubble with class
  \texttt{c("temporal\_cubble\_df",\ "cubble\_df")}
\end{itemize}

In a spatial cubble, spaital variables are organised as columns and
temporal variables are nested within a specialised \texttt{ts} column.
The cubble object printed below contains weather records in three
airport stations (the creation of a cubble object will be explained in
Section @ref(create)). This toy data is a subset of a larger data
\code{climate_aus} collected from Global Historical Climatology Network
Daily (GHCND). It records three airport stations located in Melbourne,
Australia and includes spatial variables such as station ID, longitude,
latitude, elevation, station name, World Meteorology Organisation ID.
The dataset contains temporal variables including precipitation, maximum
and minimum temperature, which can be read from the cubble header:

\begin{CodeChunk}
\begin{CodeInput}
R> cb_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form
# spatial:  [144.8321, -37.98, 145.0964, -37.6655], Missing CRS!
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name              wmo_id ts               
  <chr>       <dbl> <dbl> <dbl> <chr>              <dbl> <list>           
1 ASN00086038  145. -37.7  78.4 essendon airport   95866 <tibble [10 x 4]>
2 ASN00086077  145. -38.0  12.1 moorabbin airport  94870 <tibble [10 x 4]>
3 ASN00086282  145. -37.7 113.  melbourne airport  94866 <tibble [10 x 4]>
\end{CodeOutput}
\end{CodeChunk}

In a temporal cubble, temporal variables are expanded in the long form
and spatial variables are stored as a data attribute:

\begin{CodeChunk}
\begin{CodeInput}
R> cb_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, long form
# temporal: 2020-01-01 -- 2020-01-10 [1D], no gaps
# spatial:  long [dbl], lat [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00086038 2020-01-01     0  26.8  11  
2 ASN00086038 2020-01-02     0  26.3  12.2
3 ASN00086038 2020-01-03     0  34.5  12.7
4 ASN00086038 2020-01-04     0  29.3  18.8
5 ASN00086038 2020-01-05    18  16.1  12.5
# i 25 more rows
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{the-cubble-attributes}{%
\subsubsection{The cubble attributes}\label{the-cubble-attributes}}

A cubble object inherits the attributes from tibble (and its
subclasses): \texttt{class}, \texttt{row.names}, and \texttt{names}.
Additionally, it has three specialised attributes: \texttt{key},
\texttt{index}, and \texttt{coords}. Readers who are familiar with the
\texttt{key} and \texttt{index} attributes from the \texttt{tsibble}
package would already understand the two arguments. In cubble, the
\texttt{key} attribute identifies the row in the spatial cubble (given
the use of \texttt{tidyr::nest()} for nesting), and when combined with
the \texttt{index} argument, it identifies the row in the temporal
cubble. Currently, cubble only supports one variable as the key, and the
accepted temporal classes for index includes the base R classes
\texttt{Date}, \texttt{POSIXlt}, \texttt{POSIXct}, as well as tsibble's
\texttt{yearmonth}, \texttt{yearweek}, and \texttt{yearquarter} classes.

The \texttt{coords} attribute represents an ordered pair of coordinates.
It can be either an unprojected pair of longitude and latitude, or a
projected easting and northing value. The \texttt{sf} package is used
under the hood to calculate the bounding box, displayed in the header of
a spatial cubble.

The temporal cubble has a special attribute called \texttt{spatial} to
store the spatial variables. Shortcut functions are available to extract
attributes, for example, \code{spatial()} for extracting spatial
variables from the temporal cubble:

\begin{CodeChunk}
\begin{CodeInput}
R> spatial(cb_long)
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 3 x 6
  id           long   lat  elev name              wmo_id
  <chr>       <dbl> <dbl> <dbl> <chr>              <dbl>
1 ASN00086038  145. -37.7  78.4 essendon airport   95866
2 ASN00086077  145. -38.0  12.1 moorabbin airport  94870
3 ASN00086282  145. -37.7 113.  melbourne airport  94866
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{create}{%
\subsection{Creation and coercion}\label{create}}

In many cases, spatio-temporal data arrive in separate tables for
analysis. For example, in climate data, analysts may initially receive
station data containing geographic location information, recorded
variables and their recording periods. They can then query the temporal
variables using the stations of interest to obtain the relevant temporal
data. Alternatively, analyses may begin as purely spatial or temporal,
and analysts may obtain additional temporal or spatial data to expand
the result to spatio-temporal.

The function \code{make_cubble()} composes a \code{cubble} object from a
spatial table (\code{spatial}) and a temporal table (\code{temporal}),
along with three attributes introduced in the subsection @ref(object):
\code{key}, \code{index}, and \code{coords}. The following code creates
the spatial \code{cubble} object shown previously:

\begin{CodeChunk}
\begin{CodeInput}
R> make_cubble(spatial = stations, temporal = meteo,
+             key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\end{CodeChunk}

Spatio-temporal data in other foreign objects can be coerced into a
\code{cubble} object with the function \code{as_cubble()}. This includes
casting from a \code{tibble} or \code{data.frame} with both spatial and
temporal information, a NetCDF object, a \code{stars} object
\citep{stars}, and a \code{sftime} object \citep{sftime}. The two
examples below show the casting from a tibble and a NetCDF object. The
dataset \code{climate_flat} combines the spatial data, \code{stations},
with the temporal data, \code{meteo}, into a single tibble object. It
can be coerced into a cubble using:

\begin{CodeChunk}
\begin{CodeInput}
R> climate_flat |> as_cubble(key = id, index = date, coords = c(long, lat))
\end{CodeInput}
\begin{CodeOutput}
# cubble:   key: id [3], index: date, nested form
# spatial:  [144.8321, -37.98, 145.0964, -37.6655], Missing CRS!
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id           long   lat  elev name              wmo_id ts               
  <chr>       <dbl> <dbl> <dbl> <chr>              <dbl> <list>           
1 ASN00086038  145. -37.7  78.4 essendon airport   95866 <tibble [10 x 4]>
2 ASN00086077  145. -38.0  12.1 moorabbin airport  94870 <tibble [10 x 4]>
3 ASN00086282  145. -37.7 113.  melbourne airport  94866 <tibble [10 x 4]>
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{functions-and-methods}{%
\subsection{Functions and methods}\label{functions-and-methods}}

Table @ref(tab:funs) summarises the functions implemented in the
\pkg{cubble} package and Table @ref(tab:methods) details the methods
implemented for each of the three cubble classes. The \code{cubble_df}
class handles methods that behave consistently in both spatial
(\code{spatial_cubble_df}) and temporal cubble
(\code{temporal_cubble_df}). When the method has distinct behavior for
the spatial cubble and temporal cubble, it is implemented separately for
each subclass.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8750}}@{}}
\caption{(\#tab:funs) Functions implemented in the \pkg{cubble}
package}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Category
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Functions
\end{minipage} \\
\midrule()
\endhead
base R & \texttt{{[}}, \texttt{{[}{[}\textless{}-},
\texttt{names\textless{}-} \\
tidyverse & \texttt{dplyr\_row\_slice}, \texttt{dplyr\_col\_modify},
\texttt{dplyr\_reconstruct}, \texttt{select}, \texttt{mutate},
\texttt{arrange}, \texttt{filter}, \texttt{group\_by}, \texttt{ungroup},
\texttt{summarise}, \texttt{select}, \texttt{slice}, \texttt{rowwise},
\texttt{rename}, \texttt{bind\_rows}, \texttt{bind\_cols},
\texttt{relocate}, \texttt{type\_sum}, the slice family
(\texttt{slice\_head}, \texttt{slice\_tail}, \texttt{slice\_max},
\texttt{slice\_min}, \texttt{slice\_sample}) and the join family
(\texttt{left\_join}, \texttt{right\_join}, \texttt{inner\_join},
\texttt{full\_join}, \texttt{anti\_join}, \texttt{semi\_join}) \\
cubble & \texttt{as\_cubble}, \texttt{cubble}, \texttt{make\_cubble},
\texttt{check\_key} \texttt{face\_temporal}, \texttt{face\_spatial},
\texttt{unfold}, \texttt{key}, \texttt{key\_vars}, \texttt{key\_data},
\texttt{index}, \texttt{index\_var}, \texttt{coords}, \texttt{spatial},
\texttt{match\_sites}, \texttt{match\_spatial},
\texttt{match\_temporal}, \texttt{geom\_glyph},
\texttt{geom\_glyph\_box}, \texttt{geom\_glyph\_line},
\texttt{make\_spatial\_sf}, \texttt{make\_temporal\_tsibble},
\texttt{fill\_gaps}, and \texttt{scan\_gaps} \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7500}}@{}}
\caption{(\#tab:methods) Methods implemented for the three \code{cubble}
classes.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Class
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Methods
\end{minipage} \\
\midrule()
\endhead
\texttt{cubble\_df} &
\texttt{{[}{[}\textless{}-,\ dplyr\_col\_modify,\ key\_data,\ key\_vars,\ key,\ print} \\
\texttt{spatial\_cubble\_df} &
\texttt{{[},\ names\textless{}-,\ \ tbl\_sum,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ arrange,\ rename,\ rowwise,\ group\_by,\ ungroup,\ select,\ spatial,\ summarise,\ unfold,\ update\_cubble} \\
\texttt{temporal\_cubble\_df} &
\texttt{{[},\ names\textless{}-,\ tbl\_sum,\ arrange,\ dplyr\_reconstruct,\ dplyr\_row\_slice,\ face\_spatial,\ face\_temporal,\ unfold,\ fill\_gaps,\ group\_by,\ ungroup,\ \ rename,\ rowwise,\ scan\_gaps,\ select,\ spatial,\ summarise,\ tbl\_sum,\ bind\_rows,\ bind\_cols,\ update\_cubble} \\
\bottomrule()
\end{longtable}

Specifically, the pair of verbs, \code{face_temporal()} and
\code{face_spatial()}, pivot the cubble object between the spatial and
temporal face of the multivariate spatio-temporal cube, as illustrated
in Figure \ref{fig:face}. The code below uses the function
\code{face_temporal()} and \code{face_spatial()} to transform between a
spatial cubble (\code{cb_nested}) and a temporal one (\code{cb_long}),
introduced earlier in subsection @ref(object):

\begin{CodeChunk}
\begin{CodeInput}
R> identical(face_temporal(cb_nested), cb_long)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> identical(face_spatial(cb_long), cb_nested)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

The pair of verbs are exact inverse and apply both functions on a cubble
object will result in the object itself:

\begin{CodeChunk}
\begin{CodeInput}
R> identical(face_spatial(face_temporal(cb_nested)), cb_nested)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> identical(face_temporal(face_spatial(cb_long)), cb_long)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

@@

{\centering \includegraphics[width=1\linewidth]{figures/diagram-keynotes/diagram-keynotes.001}

@@

\caption{An illustration of the function \code{face\_temporal()} and \code{face\_spatial()}: \code{face\_temporal()} converts a spatial cubble (neseted form) into a temporal cubble (long form) to focus on the temporal variables. Conversely, \code{face\_spatial()} transforms a temporal cubble into a spatial one to emphasize spatial variables.}(\#fig:face)
\end{figure}
\end{CodeChunk}

\hypertarget{compatibility-with-tsibble-and-sf}{%
\subsection{Compatibility with tsibble and
sf}\label{compatibility-with-tsibble-and-sf}}

Analysts often have their preferred spatial or temporal data structure
that they prefer to use for spatio-temporal analysis. For example, the
\code{tbl_ts} class from the \code{tsibble} package \citep{tsibble} is
commonly used in time series forecasting and similarly, the \code{sf}
class \citep{sf} is frequently used in spatial data science. In
\code{cubble}, analysts have the flexibility to combine these two
structures together by allowing the spatial component to also be an
\code{sf} object and the temporal component to also be a \code{tsibble}
object.

\hypertarget{using-a-tsibble-object-as-the-temporal-component}{%
\subsubsection{Using a tsibble object as the temporal
component}\label{using-a-tsibble-object-as-the-temporal-component}}

The \code{key} and \code{index} arguments in a \code{cubble} object
corresponds to the \code{tsibble} counterparts and they can be safely
omitted, if the temporal component is a \code{tsibble} object,
i.e.~\code{meteo_ts} in the example below. The \code{tsibble} class from
the input will be carried over to the \code{cubble} object:
